---
title: "WRS Stream Temp Downlaod from USGS-NWIS"
format: html
editor: visual
author: "Nathan Lau & Hayden Waters"
date: "4/30/26"
---

##Load packages

```{r}
#install.packages("dataRetrieval")
library(dataRetrieval)
library(tidyverse)
library(here)
```

##Create a data frame

Data frame for USGS stream temperature data from the Willamette River in Oregon. Data will be for three sites from 01/01/2010 to 12/31/2023.

```{r}
siteNumbers <- c("14211720", "14211010", "14166000")
parameterCd <- "00010"  # water temperature
startDate <- "2010-01-01"
endDate <- "2023-12-31"

temp_data <- readNWISdv(siteNumbers, parameterCd, startDate, endDate)
```

```{r}
# rename columns for clarity
colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_C")
# extract some date info for use later
temp_data$month = month(temp_data$Date, label=TRUE)
temp_data$year = year(temp_data$Date)

# check by plotting
ggplot(temp_data, aes(x=Date, y=temp_C, color=site_no)) +
  geom_line() +
  labs(title="Stream Temperature at Willamette River Sites",
       x="Date", y="Temperature (°C)", color="Site Number") +
  theme_minimal()
```

##Create a CSV of the downloaded stream data

```{r}
write.csv(temp_data, file = here("HW2", "wrs_stream_temp.csv"))
```

##Download data from CSV

```{r}
temp_data_download <- read.csv(here("HW2", "wrs_stream_temp.csv"))|>
  select(-X, -NA.) |>
  mutate(site_no = as.factor(site_no))
```

##Create a boxplot of stream temperature

```{r}
boxplot_1 <- ggplot(temp_data_download, aes(site_no, temp_C))+
  geom_boxplot() +
  labs(title = "Stream Temperature at Willamette River Sites",
       x = "Site", y = "Temperature (°C)")
boxplot_1
```

Figure 1. Stream temperature in °C is plotted for the Willamette River at three sites.

##Create data frame of summary stats by site Summary stats calculated: mean, minimum, maximum, and standard deviation values by site.

```{r}
wrs_summary_stats <- temp_data_download |>
  group_by(site_no) |>
  summarize(mean_temp = mean(temp_C), 
            min_temp = min(temp_C), 
            max_temp = max(temp_C),
            std_temp = sd(temp_C))
```

##Calculate the trend in mean April stream temperatures for each site from year to year by fitting a Generalized Linear Model (GLM)

```{r}
# filter for April temperatures
april_data <- subset(temp_data_download, month == "Apr")
# fit a GLM model to see if there is a trend in April temperatures over time, separated by site
april_model <- lm(temp_C ~ year*site_no, data = april_data)
summary(april_model)
```

The LM model was statistically significant (p < 0.05) but has an r-squared of 0.32, meaning only 32% of the variance with site and year is explained by the model. While the linear found an increase of 0.025963 degrees C/year for site 14166000, an increase of 0.023099 degrees C/year for site 14211010, and an increase of 0.032797 degrees C/year for site 14211720, none of these trends are signficant (p > 0.05).

