---
title: "WRS Stream Temp Downlaod from USGS-NWIS"
format: html
editor: visual
author: "Nathan Lau & Hayden Waters"
date: "4/30/26"
---

##Load packages

```{r}
#install.packages("dataRetrieval")
library(dataRetrieval)
library(tidyverse)
library(here)
```

##Create a data frame

Data frame for USGS stream temperature data from the Willamette River in Oregon. Data will be for three sites from 01/01/2010 to 12/31/2023.

```{r}
siteNumbers <- c("14211720", "14211010", "14166000")
parameterCd <- "00010"  # water temperature
startDate <- "2010-01-01"
endDate <- "2023-12-31"

temp_data <- readNWISdv(siteNumbers, parameterCd, startDate, endDate) #create a data frame of stream data
```

```{r}
# rename columns for clarity
colnames(temp_data) <- c("agency_cd", "site_no", "Date", "temp_C")
# extract some date info for use later
temp_data$month = month(temp_data$Date, label=TRUE)
temp_data$year = year(temp_data$Date)

# check by plotting
ggplot(temp_data, aes(x=Date, y=temp_C, color=site_no)) +
  geom_line() +
  labs(title="Stream Temperature at Willamette River Sites",
       x="Date", y="Temperature (°C)", color="Site Number") +
  theme_minimal()
```

##Create a CSV of the downloaded stream data

```{r}
write.csv(temp_data, file = here("HW2", "wrs_stream_temp.csv"))
```

##Download data from CSV

```{r}
temp_data_download <- read.csv(here("HW2", "wrs_stream_temp.csv"))|>
  select(-X, -NA.) |> #removing extra columns
  mutate(site_no = as.factor(site_no)) #site_no as factor for boxplots by site number
```

##Create a boxplot of stream temperature

```{r}
boxplot_1 <- ggplot(temp_data_download, aes(site_no, temp_C))+
  geom_boxplot() +
  labs(title = "Stream Temperature at Willamette River Sites",
       x = "Site", y = "Temperature (°C)")
boxplot_1
```

Figure 1. Stream temperature in °C is plotted for the Willamete River at three sites.

##Create data frame of summary stats by site Summary

```{r}
# Calculate stats: mean, minimum, maximum, and standard deviation by site
wrs_summary_stats <- temp_data_download |>
  group_by(site_no) |>
  summarize(mean_temp = mean(temp_C), 
            min_temp = min(temp_C), 
            max_temp = max(temp_C),
            std_temp = sd(temp_C))
```

##October temperature analysis

```{r}
october_temp <- subset(temp_data_download, month == "Oct") # Filter downloaded data for October
october_model <- lm(temp_C ~ year * site_no, data = october_temp) #Linear model
summary(october_model)

boxplot_2 <- ggplot(october_temp, aes(site_no, temp_C))+
  geom_boxplot() +
  labs(title = "October Stream Temperature at Willamette River Sites",
       x = "Site", y = "Temperature (°C)")
boxplot_2
```

The linear model is well-fitted to the October data with a p-value < 0.05. The r-squared of 0.26, indicates that 26% of the temperature variance is explained by the model. The year slope estimate is not significant (p > 0.05), meaning the change in temperature over time is not different from random chance. The water temperatures between the three sites were not significantly different. The same is true when the year-site interaction is included.
